Missing the Point: Non-Convergence in Iterative Imputation Algorithms

### 2020-04-27 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
After textual feedback Gerko

\begin{abstract}
(\textbf{Rewrite:}) Multiple imputation by chained equations (MICE) is a
widely used tool to accommodate missing data. While empirical evidence
supports the validity of inferences obtained using mice, there is no
consensus on the convergence properties of the method. This paper
provides insight into non-convergence of mice algorithms.
\end{abstract}

@@

Anyone who analyzes person-data may run into a missing data problem.
Missing data is not only ubiquitous but also tedious. If a dataset
contains just one incomplete observation, statistical inferences are
undefined and will not produce any results. To circumvent this, many
statistical packages employ list-wise deletion by default (i.e.,
ignoring incomplete observations). Unfortunately, this \emph{ad hoc}
solution may yield wildly invalid results \citep{buur18}. An alternative
is to `impute' (i.e., fill in) the missing values in the incomplete
observations. Subsequently, statistical inferences can be performed on
the completed dataset. By repeating this process several times, a
distribution of plausible results may be obtained, which reflects the
uncertainty in the data due to missingness. This technique is known as
`multiple imputation' \citep[MI;][]{rubin76}. MI yields unbiased and
confidence valid estimates of the true---but missing---data inference
under many circumstances \citep{buur18}.

\textbf{Move this after terminology??} Figure \ref{fig:diagram} provides
an overview of the steps involved with MI---from incomplete data, to
\(M\) completed datasets, to \(M\) estimated quantities of interest
(\(\hat{Q}\)s), to a single pooled estimate \(\bar{Q}\). Missing data in
\(y\) is `imputed' (i.e., filled in) \(M\) times. The imputed data
(\(y_{imp}\)) is combined with the observed data (\(y_{obs}\)) to create
\(M\) completed datasets. On each completed dataset, the analysis of
scientific interest (the `substantive model', or `complete data model')
is performed. The quantity of scientific interest (e.g., a regression
coefficient) is denoted with \(Q\). Since \(Q\) is estimated on each
completed dataset, \(M\) separate \(\hat{Q}\)-values are obtained. These
\(M\) values are combined into a single pooled estimate \(\bar{Q}\).
\textbf{ALSO: make clear that iteration happens within the first step,
between incomplete and completed data.}

@@

A popular method to obtain imputations is to use the `Multiple
Imputation by Chained Equations' algorithm, shorthand
`MICE'\citep{mice}. MICE is an iterative algorithmic procedure to draw
imputations from the posterior predictive distribution of the missing
values. This introduces a potential threat to the validity of the
imputations: What if the algorithm has not converged? Are the
implications then to be trusted? And can we rely on the inference
obtained on the completed data? These are all open questions, because
the convergence properties of iterative imputation algorithms have not
been systematically studied \citep{buur18}. Moreover, there is no
scientific consensus on the convergence properties of MI algorithms
\citep{taka17}. Some default MICE techniques (e.g., `predictive mean
modeling') might not yield converged states at all \citep{murr18}.
Therefore, algorithmic convergence should be monitored carefully.

@@

In this paper, we investigate how non-convergence in iterative
imputation algorithms may be diagnosed, how well these methods perform,
and at which point convergence may safely be assumed. For reasons of
brevity, we only focus on the MI algorithm implemented in the
world-leading MI software: the \texttt{R} \citep{R} package
\texttt{mice} \citep{mice}. The convergence properties of this MI
algorithm are investigated through model-based simulation. The results
of this simulation study are guidelines for assessing convergence of MI
algorithms, which will aid applied researchers in drawing valid
inference from incomplete datasets.

\textbf{Include sub-questions?:} How can non-convergence be diagnosed?
Are common MCMC non-convergence diagnostics appropriate for MICE? And if
so, which threshold should be used to diagnose non-convergence? How many
iterations are sufficient/needed to be able to diagnose non-convergence?
Are the default number of iterations sufficient (i.e., 5 in mice, 10 in
SPSS and Stata, 30 in mi)?

\hypertarget{terminology}{%
\subsection{Terminology}\label{terminology}}

@@

\hypertarget{how-may-non-convergence-be-diagnosed}{%
\section{How may non-convergence be
diagnosed?}\label{how-may-non-convergence-be-diagnosed}}

The two challenges for convergence of iterative algorithms are mixing
and stationarity \citep{gelm13}. Without mixing, imputation chains do
not intermingle nicely. Without stationarity, there is trending within
imputation chains, see Figure \ref{fig:non-conv}. \textbf{Explain what
we see, namely example by \citet{buur18} reproduced, showing the
traceplots of chain means for some variable. The first plot is typical
convergence of MICE, the second is pathological non-convergence because
of a mis-specified imputation model. Each line is an imputation. In the
first plot, the lines intermingle nicely and there is little trending.
In the second, there is a lot of trending and the lines do not cross
much. Importantly, the chain means at the last iteration (the imputed
value per \(m\)) are very different between the two plots. The algorithm
with the mis-specified model yields imputed values that are on average a
magnitude two larger than those of the typically converged algorithm.
This shows the importance of reaching converged states in iterative
imputation algorithms.}

Non-stationarity within chains may be diagnosed with e.g.,
autocorrelation \citep[\(AC\);][]{scha97, gelm13}, numeric standard
error \citep[`MC error';][]{gewe92}, or Raftery and Lewis's
\citeyearpar{raft91} procedure to determine the effect of trending on
the precision of estimates. A widely used diagnostic to monitor mixing
between chains is the potential scale reduction factor \(\widehat{R}\)
\citep[`Gelman-Rubin statistic';][]{gelm92}. With a recently proposed
adaptation, \(\widehat{R}\) might also serve to diagnose
non-stationarity, but this has not yet been thoroughly investigated
\citep{veht19}. Therefore, mixing and stationarity will be evaluated
separately in this study. As recommended \citep[e.g.,][p.~898]{cowl96},
\(AC\) and \(\widehat{R}\) will be used.

@@

From the between- and within-chain variances we compute a weighted
average, \(\widehat{\operatorname{var}}^{+}\), which over-estimates the
total variance of \(\theta\) \textbf{remove or explain why}.
\(\widehat{R}\) is then obtained as a ratio between the over-estimated
total variance and the within-chain variance:

@@

We can interpret \(AC\)-values as a measure of stationarity. If
\(AC\)-values are close to zero, there is no dependence between
subsequent samples within imputation chains. Negative \(AC\)-values
indicate divergence within imputation chains. \textbf{Subsequent sampled
values within each imputation chain are less alike.} Positive
\(AC\)-values indicate recurrence. If \(\theta\)-values of subsequent
iterations are similar, trending may occur. Negative \(AC\)-values show
no threat to the stationarity component of convergence. On the contrary
even---negative \(AC\)-values indicate that \(\theta\)-values of
subsequent iterations diverge from one another, which may increase the
variance of \(\theta\) and speed up convergence. As convergence
diagnostic, the interest is therefore in positive \(AC\)-values.
\textbf{Maybe remove:} Moreover, the magnitude of \(AC\)-values may be
evaluated statistically, but that is outside of this note's scope.

@@

\textbf{To assess whether \(\widehat{R}\) and \(AC\) may be appropriate
non-convergence diagnostics for MICE algorithms, the following condition
must hold when applied on the two algorithms plotted above: the methods
indicate worse performance for the mis-specified model with pathological
non-convergence (i.e., higher \(\widehat{R}\)- and \(AC\)-values than
the typical performance). And additionally the methods should reflect
the increasing convergence in the typical convergence situation as the
number of iterations goes up.}

\textbf{In figure \ref{fig:diagnostics}A (}add panel labels\textbf{),
the chain means are plotted again, now together. So are the
non-convergence diagnostics. Panel B shows \(\widehat{R}\) as computed
by implementing \citet{veht19} 's recommendations. As required,
\(\widehat{R}\) indicates less signs of non-convergence as the number of
iterations goes up in the typical convergence situation. The superior
performance of the typical convergence over the pathological
non-convergence is less prominent, even flipped for \(t<4\). Panel C
displays the \(AC\) as computed with the R function stats::acf(). When
we look at this panel, we must conclude that something weird is going
on. The \(AC\)-values indicate equal performance (up-to \(t=5\)) for the
typical convergence and the pathological non-convergence, while there is
obvious trending in the latter. Moreover, the best convergence (as
indicated by the lowest \(AC\)-value) is observed at \(t=2\), but
looking at the chain means in panel A, there should be some signs of
trending up-to iteration number seven. After consulting the
documentation on stats::acf(), we conclude that this implementation of
\(AC\) is not suitable for iterative imputation algorithms. There is a
correction factor for a mathematical shortcut that works in the limit.
According to\ldots{} the function works when the number of iterations
\textgreater{} 50. The default number of iterations in mice, however, is
5. Therefore, we compute \(AC\) manually, see panel D. The AC values in
this plot do meet the requirements (\ldots) and will therefore be used
in the simulation study.}

@@

\caption{Convergence diagnostics pathological non-convergence versus typical convergence.}\label{fig:diagnostics}
\end{figure}

@@

Convergence is reached when there is no dependency between subsequent
iterations of imputation chains (\(AC = 0\)), and chains intermingle
such that the only difference between the chains is caused by the
randomness induced by the algorithm (\(\widehat{R} = 1\)). We expect
that a completely converged state will not be reached.

This study evaluates whether \(\widehat{R}\) and \(AC\) could diagnose
convergence of multiple imputation algorithms. We assess the performance
of the two convergence diagnostics against the recommended evaluation
criteria for MI methods \citep[i.e., average bias, average confidence
interval width, and empirical coverage rate across simulations;][\(\S\)
2.5.2]{buur18}. That is, there is no baseline measure available to
evaluate performance against.

Based on an empirical finding \citep{lace07}, we hypothesize that
\(\widehat{R}\) will over-estimate non-convergence of MI algorithms
\textbf{explain why?}. The threshold of \(\widehat{R} < 1.01\) will then
be too stringent for diagnosing convergence. This over-estimation may,
however, be diminished because \(\widehat{R}\) can falsely diagnose
convergence if initial values of the algorithm are not appropriately
over-dispersed \citep[p.~437]{broo98}. In \texttt{mice}, initial values
are chosen randomly from the observed data. Therefore, we cannot be
certain that the initial values are over-dispersed. We expect this to
have little effect on the hypothesized performance of \(\widehat{R}\)
\textbf{explain that the randomness induced by the MICE algorithm would
take care of this??}. \textbf{(Add actual hypothesis: we thought that
mice would converge sooner than \(\widehat{R}\) would indicate that it
did)} No hypothesis was formulated about the performance of \(AC\) as
convergence diagnostic. \textbf{Added later: High \(AC\)-values are
implausible in MI procedures. That is, the randomness induced by the MI
algorithm effectively mitigates the risk of dependency within chains.
But this doesn't seem the case in practice??}

@@

The two challenges of convergence are stationarity and mixing. Mixing
and stationarity can be inspected visually, by evaluating traceplots of
scalar summaries of interest (\(\theta\)s; e.g., chain means and chain
variances). As \citet{buur18} describes, users can also specify a
model-specific scalar summary (e.g., a regression coefficient). A
user-specified scalar summary, however, is not universal to all complete
data problems. Therefore, as inspired by \citep{mack03}, we will propose
a \(\theta\) that summarizes the multivariate state space of the
algorithm. Namely, the first eigenvalue of the variance-covariance
matrix of the \(M\) completed datasets.

@@

The performance of two methods (\(\widehat{R}\) and \(AC\)) on each of
these scalar summaries of interest will be evaluated with several
estimands (quantities of scientific interest; \(Q\)s). It is assumed
that when \(\bar{Q}\) (the pooled result across imputations) is an
unbiased (and confidence valid) estimate of \(Q\), the algorithm is
sufficiently converged. For each scalar \(\theta\), one estimand \(Q\)
is defined that may be of interest in empirical research. The
\(\widehat{R}\) and \(AC\) values with chain means as \(\theta\)s are
evaluated against the bias in univariate mean estimates (\textbf{all
variables or just for Y??}). Similarly, the performance measure for
\(\widehat{R}\) and \(AC\) applied to chain variances is the bias in
estimated standard deviations. To evaluate the performance of
\(\widehat{R}\) and \(AC\) on the eigenvalues, we will use bias in the
coefficient of determination, \(R^2\). Additionally, we will evaluate
the estimated regression coefficients, the coverage rate of the CI95\%
of the regression estimates, and the CI length (\textbf{see definitions
in old Methods}). \textbf{remove the term performance measure here and
add an extra paragraph to define performance measures as bias in all
estimands, and coverage rate and CI length of regression coefficients.}

@@

Upon convergence, \(\widehat{R}=1\) and \(AC=0\), but in MCMC algorithms
these are very unlikely (because it's a distribution, not a point to
converge to). In practice, thresholds are used to diagnose
non-convergence. Usually, non-convergence is diagnosed when
\(\widehat{R}\) \textgreater{} 1.2 or 1.1 or even 1.01. And a t-test is
performed to assess whether \(AC\) is significantly different from zero.
Performance measures are the same as above: unbiased, confidence valid
estimates.

\textbf{Move this to Results:} As expected, complete convergence
(\(\widehat{R}=1\) and \(AC=0\)) is not observed. But what about the
thresholds for practice? \(\widehat{R}\) \textless1.2 is not stringent
enough, because conditions in which \(\widehat{R}\) is smaller than 1.2
do not all yield unbiased estimates of the \(Q\)s. Moreover, there is a
dip in the \(\widehat{R}\) values, after which \(\widehat{R}\)
\textgreater{} 1.2 again. If the algorithms would be terminated at the
iteration where \(\widehat{R}\) \textless{} 1.2 occurs for the first
time, the increase in \(\widehat{R}\) values might be missed. The
threshold \(\widehat{R}\) \textless{} 1.1 is somewhat conservative in
comparison with the performance measures. But it seems OK.
\(\widehat{R}\) \textless{} 1.01 is too stringent compared to the
performance measures. And it is not obtained for any number of
iterations (\textbf{but may be necessary in more complex complete data
models???}).

@@

\begin{verbatim}
# pseudo-code of simulation 
1. simulate data 
for (missingness proportions 5%, 25%, 50%, 75% and 95%)
 for (number of simulation runs from 1 to 1000)
  2. create missingness
  for (number of iterations from 1 to 100)
   3. impute missingness
   4. compute convergence diagnostics
   5. perform analysis of scientific interest
   6. pool results across imputations
   7. compute performance measures
 8. aggregate outcomes across simulation runs 
9. combine outcomes of all missingness proportions
\end{verbatim}

@@

The data-generating mechanism is a multivariate normal distribution,
representing person data on three independent variables (from an
unspecified social scientific field of study). Let

\begin{align*}
\begin{pmatrix}X_1\\
X_2\\
X_3
\end{pmatrix} \sim \mathcal{N}
\begin{bmatrix}
\begin{pmatrix}
12\\
3\\
0.5
\end{pmatrix}\!\!,
\begin{pmatrix}
4 & 4 & 1.8 & 0\\
4 & 16 & 4.8 & 0\\
1.8 & 4.8 & 9 & 0
\end{pmatrix}
\end{bmatrix}\!\!\text{.}\\[2\jot]
\end{align*} In this study, sampling variance is not of interest.
Therefore, a single complete set may serve as comparative truth in all
simulation runs \citep{vinknd}. A finite population of \(N=1000\) is
simulated using the function \texttt{mvtnorm::rmvnorm()}. Subsequently,
a fourth variable is constructed to serve as dependent variable in a
multiple linear regression problem. Let \[
Y_i = 1 + 2X_{1i} +.5X_{2i} - X_{3i} + \epsilon_i ,
\] where \(i = 1, 2,..., N\) and \(\epsilon \sim \mathcal{N}(0, 100)\).

We consider four quantities of scientific interest \citep[`conceptual
estimands';][]{morr19}, namely\ldots{} solve a multiple linear
regression problem, where dependent variable \(Y\) is regressed on
independent variables \(X_1\), \(X_2\) and \(X_3\) (\textbf{check
notation with betas here, and add what the quantity/-ies of scientific
interest is/are! Namely, the descriptive statistics of all variables
(mean and standard deviation), \(\beta_2\), and the percentage of
variance in \(Y\) explained (coefficient of determination; \(R^2\))}):

@@

The complete data is `amputed' once for each simulation repetition with
function \texttt{mice::ampute()}. (\textbf{change this to reflect
current simulation set-up with 5, 25, 50, 75, and 95\% of cases having
missing data:} The missingness is univariate, and the probability to be
missing is the same for all four variables, namely 20\%
(\texttt{prop\ =\ 0.8,\ mech\ =\ "MCAR"}). This leaves 20\% of the rows
completely observed). This study only considers only an MCAR missingness
mechanism. Therefore, results may not be extrapolated to other missing
data problems. Proper performance of the convergence diagnostics under
MCAR is necessary but not sufficient to demonstrate the appropriateness
of \(\widehat{R}\) and \(AC\) as convergence diagnostics. Convergence
diagnostics should therefore at least apply to the MCAR situation,
before more complex missingness should be explored.

Missing datapoints in \(y\) are imputed with the function
\texttt{mice::mice()}. All MI procedures are performed with Bayesian
linear regression imputation (\texttt{method\ =\ "norm"}), and five
imputation chains (\texttt{m\ =\ 5}). The number of iterations varies
between simulation conditions (\texttt{maxit\ =} \(1, 2, \dots, 100\)).
Each repetition of the simulation thus starts from 1 complete dataset,
then splits into 5 missingness conditions (\(p =.05,.25,.5,.75,.95\)),
each of the amputed datasets is imputed 5 times (\(m = 1, 2,..., 5\))
for each iteration condition (\(t = 1, 2,..., 100\)). From these
\(5\times5\times100\) imputations, we extract the \(\theta\)s to apply
\(\widehat{R}\) and \(AC\) on (\textbf{i.e., chain means, chain
variances, \(\beta\)s per imputation, and the first eigenvalue of the
variance-covariance matrix per imputation}). We calculate
\(\widehat{R}\) by implementing Vehtari et al.'s \citeyearpar{veht19}
recommendations, and \(AC\) as the correlation between the \(t^{th}\)
and the \((t+1)^{th}\) iteration. For each \(\theta\) we estimate the
corresponding scientific estimand \(Q\).

>>>>

\begin{abstract}
(\textbf{Rewrite:}) Iterative imputation by chained equations (MICE) is
a widely used tool to accommodate missing data. While it is widely
accepted that valid inferences can be obtained with this technique,
these inferences all rely on algorithmic convergence. There is no
consensus on how to evaluate the convergence properties of the method.
This paper provides insight into identifying non-convergence of
iterative impuation algorithms.
\end{abstract}

@@

Anyone who analyzes person-data may run into a missing data problem.
Missing data is not only ubiquitous but also \textbf{tedious}. If a
dataset contains just one incomplete observation, statistical inferences
are undefined and will not produce any results. To circumvent this, many
statistical packages employ list-wise deletion by default (i.e.,
ignoring incomplete observations). Unfortunately, this \emph{ad hoc}
solution may yield wildly invalid results \citep{buur18}. An alternative
is to `impute' (i.e., fill in) the missing values in the incomplete
observations. Subsequently, statistical inferences can be performed on
the completed dataset. By repeating this process several times, a
distribution of plausible results may be obtained, which reflects the
uncertainty in the data due to missingness. This technique is known as
`multiple imputation' \citep[MI;][]{rubin76}. MI has proven to be a
powerful technique to yield unbiased and confidence valid estimates of
the true---but missing---data inference under many circumstances
\citep{buur18}.

\textbf{Move this after terminology??} Figure \ref{fig:diagram} provides
an overview of the steps involved with MI---from incomplete data, to
\(M\) completed datasets, to \(M\) estimated quantities of interest
(\(\hat{Q}\)s), to a single pooled estimate \(\bar{Q}\). Missing data in
\(y\) is `imputed' (i.e., filled in) \(M\) times. The imputed data
(\(y_{imp}\)) is combined with the observed data (\(y_{obs}\)) to create
\(M\) completed datasets. On each completed dataset, the analysis of
scientific interest is performed. The quantity of scientific interest
(e.g., a regression coefficient) is denoted with \(Q\). Since \(Q\) is
estimated on each completed dataset, \(M\) separate \(\hat{Q}\)-values
are obtained. These \(M\) values are combined into a single pooled
estimate \(\bar{Q}\).

@@

A popular method to obtain imputations is to use the `Multiple
Imputation by Chained Equations' algorithm, shorthand
`MICE'\citep{mice}. MICE is an iterative algorithmic procedure to draw
imputations from the posterior predictive distribution of the missing
values. This introduces a potential threat to the validity of the
imputations: What if the algorithm has not converged? Are the
implications then to be trusted? And can we rely on the inference
obtained on the completed data? These are all open questions, because
the convergence properties of iterative imputation algorithms have not
been systematically studied \citep{buur18}. Moreover, there is no
scientific consensus on how to evaluate convergence of MI algorithms
\citep{taka17}. Some default MICE techniques (e.g., `predictive mean
modeling') might not yield converged states at all \citep{murr18}.
Therefore, algorithmic convergence should be monitored carefully.

@@

In this paper, we investigate how non-convergence in iterative
imputation algorithms may be diagnosed, how well these methods perform,
and at which point convergence may safely be assumed. For reasons of
brevity, we only focus on the MI algorithm implemented in the popular
\texttt{mice} package \citep{mice} in \texttt{R} \citep{R}. The
convergence properties of the MICE algorithm are investigated through
model-based simulation. The results of this simulation study are
guidelines for assessing convergence of MI algorithms, which will aid
applied researchers in drawing valid inference from incomplete datasets.

\textbf{Include sub-questions?:} How can non-convergence be diagnosed?
Are common MCMC non-convergence diagnostics appropriate for MICE? And if
so, which threshold should be used to diagnose non-convergence? How many
iterations are sufficient/needed to be able to diagnose non-convergence?
Are the default number of iterations sufficient (i.e., 5 in mice, 10 in
SPSS and Stata, 30 in mi)? \textbf{How severe is it when the algorithm
has not converged? And what are guidelines for practice?}

\hypertarget{terminologynotation}{%
\subsection{Terminology/Notation}\label{terminologynotation}}

@@

\hypertarget{identifying-non-convergence}{%
\section{Identifying
non-convergence}\label{identifying-non-convergence}}

There are two requirements for convergence of iterative algorithms:
mixing and stationarity \citep{gelm13}. Without mixing, imputation
chains do not intermingle nicely, \textbf{indicating that \ldots{}}.
Without stationarity, there is trending within imputation chains, which
implies that further iterations would yiled a different set of
imputations. In Figure \ref{fig:non-conv}\ldots{} \textbf{Explain what
we see, namely example by \citet{buur18} reproduced, showing the
traceplots of chain means for some variable. The first plot is typical
convergence of MICE, the second is pathological non-convergence because
of a mis-specified imputation model. Each line is an imputation. In the
first plot, the chains intermingle nicely and there is little to no
trending. In the second plot, there is a lot of trending and some chains
do not intermingle. Importantly, the chain means at the last iteration
(the imputed value per \(m\)) are very different between the two plots.
The algorithm with the mis-specified model yields imputed values that
are on average a magnitude two larger than those of the typically
converged algorithm. This shows the importance of reaching converged
states in iterative imputation algorithms.}

Non-stationarity within chains may be diagnosed with e.g.,
autocorrelation \citep[\(AC\);][]{scha97, gelm13}, numeric standard
error \citep[`MC error';][]{gewe92}, or Raftery and Lewis's
\citeyearpar{raft91} procedure to determine the effect of trending on
the precision of estimates. A widely used diagnostic to monitor mixing
between chains is the potential scale reduction factor \(\widehat{R}\)
\citep[`Gelman-Rubin statistic';][]{gelm92}. With a recently proposed
adaptation, \(\widehat{R}\) might also serve to diagnose
non-stationarity, but this has not yet been thoroughly investigated
\citep{veht19}. Therefore, mixing and stationarity will be evaluated
separately in this study. As recommended \citep[e.g.,][p.~898]{cowl96},
\(AC\) and \(\widehat{R}\) will be used \textbf{to diagnose
non-convergence}.

@@

From the between- and within-chain variances we compute a weighted
average, \(\widehat{\operatorname{var}}^{+}\), which over-estimates the
total variance of \(\theta\) \textbf{remove or explain why, or leave
in}. \(\widehat{R}\) is then obtained as a ratio between the
over-estimated total variance and the within-chain variance:

@@

We can interpret \(AC\)-values as a measure of stationarity. If
\(AC\)-values are close to zero, there is no dependence between
subsequent samples within imputation chains. Negative \(AC\)-values
indicate divergence within imputation chains. \textbf{Subsequent sampled
values within each imputation chain are less alike.} Positive
\(AC\)-values indicate recurrence. If \(\theta\)-values of subsequent
iterations are similar, trending may occur. Negative \(AC\)-values show
no threat to the stationarity component of convergence. On the contrary
even---negative \(AC\)-values indicate that \(\theta\)-values of
subsequent iterations diverge from one another, which may increase the
variance of \(\theta\) and speed up convergence. As convergence
diagnostic, the interest is therefore in positive \(AC\)-values.
\textbf{Maybe remove:} Moreover, the magnitude of \(AC\)-values may be
evaluated statistically, but that is outside of our scope.

@@

\textbf{To assess whether \(\widehat{R}\) and \(AC\) may be appropriate
non-convergence identifiers for iterative imputation algorithms, the
following condition must hold when applied on the two algorithms plotted
above: the methods indicate worse performance for the mis-specified
model with pathological non-convergence (i.e., higher \(\widehat{R}\)-
and \(AC\)-values than the typical performance). And additionally the
methods should reflect the increasing convergence in the typical
convergence situation as the number of iterations goes up.}

\textbf{In figure \ref{fig:diagnostics}A (}add panel labels\textbf{),
the chain means are plotted again, now together---as are the
non-convergence diagnostics. Panel B shows \(\widehat{R}\) as computed
by implementing \citet{veht19} 's recommendations. As required,
\(\widehat{R}\) indicates less signs of non-convergence as the number of
iterations goes up }in the typical convergence situation\textbf{. The
superior performance of the typical convergence over the pathological
non-convergence is less prominent, even flipped for \(t<4\). Panel C
displays the \(AC\) as computed with the R function
\texttt{stats::acf()}. When we look at this panel, we conclude something
weird. The \(AC\)-values indicate equal performance (up-to \(t=5\)) for
the typical convergence and the pathological non-convergence, while
there is obvious trending in the latter. Moreover, the best convergence
(as indicated by the lowest \(AC\)-value) is observed at \(t=2\), but
looking at the chain means in panel A, there should be some signs of
trending up-to iteration number seven. After consulting the
documentation on stats::acf() \citep{R}, we conclude that this
implementation of \(AC\) is not suitable for iterative imputation
algorithms. There is a correction factor for a mathematical shortcut
that works in the limit. According to \citet{box15}, the function works
when the number of iterations \textgreater{} 50. The default number of
iterations in iterative imputation, however, is often lower. Therefore,
we compute \(AC\) manually, see panel D. The AC values in this plot do
meet the requirements (\ldots) and will therefore be used in the
simulation study.}

@@

\caption{Convergence diagnostics pathological non-convergence versus typical convergence. **ADD PANEL LABELS**}\label{fig:diagnostics}
\end{figure}

@@

For multiple imputation algorithms, it holds that convergence is reached
when there is no dependency between subsequent iterations of imputation
chains (\(AC = 0\)), and chains intermingle such that the only
difference between the chains is caused by the randomness induced by the
algorithm (\(\widehat{R} = 1\)). \textbf{We expect that a completely
converged state will not be reached.}

This study evaluates whether \(\widehat{R}\) and \(AC\) could diagnose
convergence of multiple imputation algorithms. We assess the performance
of the two convergence diagnostics against the recommended evaluation
criteria for MI methods \citep[i.e., average bias, average confidence
interval width, and empirical coverage rate across simulations;][\(\S\)
2.5.2]{buur18}. \textbf{That is, there is no baseline measure available
to evaluate performance against.}

Based on an empirical finding \citep{lace07}, we hypothesize that
\(\widehat{R}\) will over-estimate non-convergence of MI algorithms
\textbf{explain why?}. The threshold of \(\widehat{R} < 1.01\) will then
be too stringent for diagnosing convergence. This over-estimation may,
however, be diminished because \(\widehat{R}\) can falsely diagnose
convergence if initial values of the algorithm are not appropriately
over-dispersed \citep[p.~437]{broo98}. In e.g.~\texttt{mice}, initial
values are chosen randomly from the observed data. Therefore, we cannot
be certain that the initial values are over-dispersed. We expect this to
have little effect on the hypothesized performance of \(\widehat{R}\)
\textbf{explain that the randomness induced by the MICE algorithm would
take care of this??}. \textbf{(Add actual hypothesis: we thought that
mice would converge sooner than \(\widehat{R}\) would indicate that it
did)} No hypothesis was formulated about the performance of \(AC\) as
convergence diagnostic. \textbf{Added later: High \(AC\)-values are
implausible in MI procedures. That is, the randomness induced by the MI
algorithm effectively mitigates the risk of dependency within chains.
But this doesn't seem the case in practice??}

@@

The two challenges of convergence are stationarity and mixing. Mixing
and stationarity can be inspected visually, by evaluating traceplots of
scalar summaries of interest (\(\theta\)s; e.g., chain means and chain
variances). As \citet{buur18} describes, users can also specify a
model-specific scalar summary (e.g., a regression coefficient). A
user-specified scalar summary, however, is not universal to all complete
data problems. Therefore, as inspired by \citep{mack03}, we will propose
a \(\theta\) that summarizes the multivariate state space of the
algorithm. Namely, the first eigenvalue of the variance-covariance
matrix of the \(M\) completed datasets. \textbf{Explain
model-dependent\ldots{}}

@@

The performance of two methods (\(\widehat{R}\) and \(AC\)) on each of
these scalar summaries of interest will be evaluated with several
estimands, i.e.~quantities of scientific interest \(Q\). It is assumed
that when \(\bar{Q}\) (the pooled result across imputations) is an
unbiased and confidence valid estimate of \(Q\), the algorithm is
sufficiently converged. For each scalar \(\theta\), one estimand \(Q\)
is defined that may be of interest in empirical research. The
\(\widehat{R}\) and \(AC\) values with chain means as \(\theta\)s are
evaluated against the bias in univariate mean estimates (\textbf{all
variables or just for Y??}). Similarly, the performance measure for
\(\widehat{R}\) and \(AC\) applied to chain variances is the bias in
estimated standard deviations. To evaluate the performance of
\(\widehat{R}\) and \(AC\) on the eigenvalues, we will use bias in the
coefficient of determination, \(R^2\). Additionally, we will evaluate
the estimated regression coefficients, the coverage rate of the CI95\%
of the regression estimates, and the CI length (\textbf{see definitions
in old Methods}). \textbf{remove the term performance measure here and
add an extra paragraph to define performance measures as bias in all
estimands, and coverage rate and CI length of regression coefficients.}

@@

Upon convergence, \(\widehat{R}=1\) and \(AC=0\), which are unlikely
thresholds for MCMC algorithms, because \textbf{it's} convergence to a
distribution. In practice, non-convergence is usually diagnosed when
\(\widehat{R}\) \textgreater{} 1.2 or 1.1 or even 1.01. And a t-test is
performed to assess whether \(AC\) is significantly different from zero.
Performance measures are the same as above: unbiased, confidence valid
estimates.

\textbf{Move this to Results:} As expected, complete convergence
(\(\widehat{R}=1\) and \(AC=0\)) is never observed. But what about the
thresholds for practice? \(\widehat{R}\) \textless1.2 is not stringent
enough, because conditions in which \(\widehat{R}\) is smaller than 1.2
do not all yield unbiased estimates of the \(Q\)s. Moreover, there is a
dip in the \(\widehat{R}\) values, after which \(\widehat{R}\)
\textgreater{} 1.2 again. If the algorithms would be terminated at the
iteration where \(\widehat{R}\) \textless{} 1.2 occurs for the first
time, the increase in \(\widehat{R}\) values might be missed. The
threshold \(\widehat{R}\) \textless{} 1.1 is somewhat conservative in
comparison with the performance measures. But it seems OK.
\(\widehat{R}\) \textless{} 1.01 is too stringent compared to the
performance measures. And it is not obtained for any number of
iterations (\textbf{but may be necessary in more complex complete data
models???}).

@@

\begin{verbatim}
# pseudo-code of simulation 
1. simulate data 
for (number of simulation runs from 1 to 1000)
 for (missingness proportions 5%, 25%, 50%, 75% and 95%)
  2. create missingness
  for (number of iterations from 1 to 100)
   3. impute missingness
   4. perform analysis of scientific interest
   5. compute convergence diagnostics 
   6. pool results across imputations
   7. compute performance measures
 8. combine outcomes of all missingness proportions
9. aggregate outcomes across simulation runs 
\end{verbatim}

@@

The data-generating mechanism is a multivariate normal distribution,
representing person data on three independent variables (from an
unspecified social scientific field of study). Let \textbf{what?}

\begin{align*}
\begin{pmatrix}X_1\\
X_2\\
X_3
\end{pmatrix} \sim \mathcal{N}
\begin{bmatrix}
\begin{pmatrix}
12\\
3\\
0.5
\end{pmatrix}\!\!,
\begin{pmatrix}
4 & 4 & 1.8 & 0\\
4 & 16 & 4.8 & 0\\
1.8 & 4.8 & 9 & 0
\end{pmatrix}
\end{bmatrix}\!\!\text{.}\\[2\jot]
\end{align*} In this study, sampling variance is not of interest.
Therefore, a single complete set may serve as comparative truth in all
simulation runs \citep{vink14}. A finite population of \(N=1000\) is
simulated using the \texttt{mvtnorm} package \citep{mvtnorm}.
Subsequently, a fourth variable is constructed to serve as dependent
variable in a multiple linear regression problem. Let \[
Y_i = 1 + 2X_{1i} +.5X_{2i} - X_{3i} + \epsilon_i ,
\] where \(i = 1, 2,..., N\) and \(\epsilon \sim \mathcal{N}(0, 100)\).

\textbf{We consider four quantities of scientific interest
\citep[`conceptual estimands';][]{morr19}, namely the descriptive
statistics of all variables (mean and standard deviation), \(\beta_2\),
and the percentage of variance in \(Y\) explained (coefficient of
determination; \(R^2\)). We solve a multiple linear regression problem,
where dependent variable \(Y\) is regressed on independent variables
\(X_1\), \(X_2\) and \(X_3\)}

@@

The complete data is \emph{amputed} once for each simulation repetition
with function \texttt{mice::ampute()}.

(\textbf{change this to reflect current simulation set-up with 5, 25,
50, 75, and 95\% of cases having missing data: The missingness is
univariate, and the probability to be missing is the same for all four
variables, namely 20\% (\texttt{prop\ =\ 0.8,\ mech\ =\ "MCAR"}). This
leaves 20\% of the rows completely observed). }

Proper performance of the convergence diagnostics under at least missing
completely at random (MCAR) missingness mechanism is necessary to
demonstrate the appropriateness of \(\widehat{R}\) and \(AC\) as
convergence diagnostics. However, results may not be extrapolated to
other missingness mechanisms. \textbf{Convergence diagnostics should
therefore at least apply to the MCAR situation, before more complex
missingness should be explored.}

Missing datapoints in \(y\) are imputed with the \texttt{mice} package
\citep{mice}. All MI procedures are performed with Bayesian linear
regression imputation (\texttt{method\ =\ "norm"}), and five imputation
chains (\texttt{m\ =\ 5}). The number of iterations varies between
simulation conditions (\texttt{maxit\ =} \(1, 2, \dots, 100\)). Each
repetition of the simulation thus starts from 1 complete dataset, then
splits into 5 missingness conditions (\(p =.05,.25,.5,.75,.95\)), each
of the amputed datasets is imputed 5 times (\(m = 1, 2,..., 5\)) for
each iteration condition (\(t = 1, 2,..., 100\)). From these
\(5\times5\times100\) imputations, we extract the \(\theta\)s to apply
\(\widehat{R}\) and \(AC\) on (\textbf{i.e., chain means, chain
variances, \(\beta\)s per imputation, and the first eigenvalue of the
variance-covariance matrix per imputation}). We calculate
\(\widehat{R}\) by implementing Vehtari et al.'s \citeyearpar{veht19}
recommendations, and \(AC\) as the correlation between the \(t^{th}\)
and the \((t+1)^{th}\) iteration. For each \(\theta\) we estimate the
corresponding scientific estimand \(Q\).
